---
title: "Chicago Divvy Case Study"
author: "Chris Fair"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/fair/Documents/Google Data Analytics Professional Certificate/8 Capstone Project/Case_Study1_BikeShare/Excel Data/2021") 
```

## Divvy Customer Usage analysis

This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: 

“In what ways do members and casual riders use Divvy bikes differently?”


Including required packages.
```{r import libraries}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(readxl) #importing excel files
library(tinytex) #knitting to PDF
```

## Step 1: Importing Data

The data for this study was made publicly available by Divvy via: 
https://divvy-tripdata.s3.amazonaws.com/index.html

Throughout this report, *membership riders* will refer to riders who have a membership, while *casual riders* will refer to riders who pay by the ride.

```{r import data}
jan_2021 <- read_excel("202101-divvy-tripdata.xlsx")
feb_2021 <- read_excel("202102-divvy-tripdata.xlsx")
march_2021 <- read_excel("202103-divvy-tripdata.xlsx")
april_2021 <- read_excel("202104-divvy-tripdata.xlsx")
may_2021 <- read_excel("202105-divvy-tripdata.xlsx")
june_2021 <- read_excel("202106-divvy-tripdata.xlsx")
july_2021 <- read_excel("202107-divvy-tripdata.xlsx")
aug_2021 <- read_excel("202108-divvy-tripdata.xlsx")
sept_2021 <- read_excel("202109-divvy-tripdata.xlsx")
oct_2021 <- read_excel("202110-divvy-tripdata.xlsx")
nov_2021 <- read_excel("202111-divvy-tripdata.xlsx")
dec_2021 <- read_excel("202112-divvy-tripdata.xlsx")
jan_2022 <- read_excel("202201-divvy-tripdata.xlsx")
```

## Step 2: Wrangling Data

Now that the data has been imported, it needs to be combined into one large dataframe. To do this, first manually check each dataframes column names to make sure they are the same, then append all the rows together.
```{r combing dataframes}
colnames(jan_2021)
colnames(feb_2021)
colnames(march_2021)
colnames(april_2021)
colnames(may_2021)
colnames(june_2021)
colnames(july_2021)
colnames(aug_2021)
colnames(sept_2021)
colnames(oct_2021)
colnames(nov_2021)
colnames(dec_2021)
colnames(jan_2022)

all_trips <- bind_rows(jan_2021, feb_2021, march_2021, april_2021,may_2021,june_2021,july_2021,aug_2021,sept_2021,oct_2021,nov_2021,dec_2021,jan_2022)
```

## Step 3: Cleaning Up Data and Preparing for Analysis

Now that all of the data is in one dataframe, it's time to inspect that data frame.
```{r inspect dataframe}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.
tail(all_trips) #See the last 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. 
```

Adding additional columns to aggregate the data by.
```{r adding columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$hour <- format(all_trips$started_at,"%H")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Adding a ride_length calculation and reformatting it to use in later calculations.
```{r ride_length}
# https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
str(all_trips)
```

The dataframe includes entries where ride_length was negative that need to be removed, as well as rides lasting longer than a day, which are likely errors due to missing bikes or locking/unlocking issues. Creating a new dataframe (v2) to remove these data.
```{r new dataframe}
all_trips_v2 <- all_trips[!(all_trips$ride_length<=0 | 
                              all_trips$ride_length>86400),]
#checking to make sure new numbers make sense
mean(all_trips$ride_length)
mean(all_trips_v2$ride_length)
```

## Step 4: Descriptive Analysis

A quick check of the summary statistics to get a feel for the data and see if everything is making sense.
```{r summary check}
# Descriptive analysis on ride_length (all figures in seconds)
summary(all_trips_v2$ride_length)

# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

# See how members vs casual rides vary by month
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$month, FUN = mean)

# Setting order for the days of week when sorting
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

## Step 5: Visualizations

Creating visuals to get a clearer picture of the data and see what suspected trends might or might not exist.
```{r visuals, echo=FALSE}
# Visualization for number of rides by day of week by rider type
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Number of Rides by Day of the Week")
  
# Save each plot 
#ggsave("number_rides_by_weekday_by_ridertype.png")

# Visualization for average duration by day of week by rider type
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Ride Duration by Day of the Week")

#ggsave("duration_by_weekday_by_ridertype.png")

# Visualization for ride duration by month by rider type
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarize(number_of_rides = n(), 
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>% 
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) + 
  geom_col(position = "dodge") + ggtitle("Average Ride Duration by Ride Start Month")

#ggsave("duration_by_month_by_ridertype.png")

# Visual for number of rides by month by rider type
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month) %>% 
  ggplot(aes(x=month,y=number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge") + ggtitle("Average Number of Rides by Ride Start Month")

#ggsave("number_rides_by_month_by_ridertype.png")

# Visual for number of rides by rider by start hour by rider type
all_trips_v2 %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, hour) %>% 
  ggplot(aes(x=hour,y=number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge") + ggtitle("Average Number of Rides by Ride Start Hour")

#ggsave("number_rides_by_hour_by_ridertype.png")

# Visual for duration of rides by rider by start hour by rider type
all_trips_v2 %>% 
  group_by(member_casual, hour) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, hour) %>% 
  ggplot(aes(x=hour,y=average_duration, fill = member_casual)) + 
  geom_col(position = "dodge") + ggtitle("Average Ride Duration by Ride Start Hour")

#ggsave('duration_by_hour_ridertype.png')

# Visual for type of bike preference by rider type
all_trips_v2 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, rideable_type) %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Number of Rides of Different Bike Types")

#ggsave('bike_preference_by_membership.png')

all_trips_v2 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, rideable_type) %>% 
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + ggtitle("Average Ride Duration of Different Bike Types")
```

## Step 6: Exporting Results

Below was used to export summary files for further analysis in Tableau.
```{r exports}
# counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + 
#                       all_trips_v2$day_of_week + all_trips_v2$month + 
#                       all_trips_v2$hour, FUN = mean)
# 
# write.csv(counts, file = '~/Documents/counts.csv')
# 
# write.csv(all_trips_v2, file = '~/Documents/all_trips_vs.csv')
# #Above file was too large for Tableau, so making a smaller one.
# tableau_data <- all_trips_v2 %>%  
#   select(-c(ended_at, end_station_name, end_station_id, end_lat, end_lng))
# 
# write.csv(tableau_data, file='~/Documents/tableau_data.csv')
```

## Conclusions

Overall, membership riders take more rides than casual riders, but casual riders have a longer average ride time. This is likely caused by the pricing structure (fee to undock a bike every ride) incentivizing casual riders to take fewer, but longer trips, whereas membership riders have no price incentive since undocking is free for them. 

Additionally, membership riders are likely to use Divvy to commute to work, while casual riders appear to use it more for leisure purposes, which could be contributing to the decreased ride duration for members compared to casual riders. This is supported by the weekday data, which shows membership riders having their highest number of rides during the week, while casual riders most popular days of use are on the weekends. Further support of this comes from inspecting the hourly number of rides, which shows two peaks for membership riders, one at 8am and one at 5pm, which would be representative of typical commute hours.

Both casual and membership ridership peak during the summer months and trail off as winter approaches, with the lowest ridership occuring during the month of February. This is primarily due to the weather in Chicago during the winter months and the large amount of snow accumulation that can make riding both uncomfortable and dangerous. It is worth noting that casual riders have a more extreme difference in rides taken between summer and winter months compared to membership riders. This is likely due to a combination of membership riders still relying on Divvy bikes to commute in the winter, as well as the sunk cost of an annual membership incentivizing membership riders to continue riding, whereas casual riders have no sunk cost.

Finally, when it comes to bike preference, both casual riders and membership riders seem to prefer classic bikes over electric bikes. This preference appears to be stronger amongst membership riders and, for both types of riders, appears to be driven by the increased cost of the electric bike compared to a classic bike.

## Recommendations

The following are just a few available options that could help convert more riders from casual riders to membership riders:

* Make a marketing pushing in the early summer to convert casual riders to membership riders as this is when Divvy is most in demand.
* Explore varying price structure on weekdays vs weekends to see if casual riders can be incentivized to convert to memberships.
* Explore allowing members to reserve bikes during certain weekend hours when bikes would be most in-demand.

Additional work could be done in the following areas to provide a clearer picture of what motivates Divvy users:

* Explore the relationship between start/end ride locations and rider type (member vs casual).
* Investigate more precisely how weather relates to members/casual riders willingness to ride (in both summer and winter months).
* Investigate how large city events (Lollapalooze, Chicago Marathon, etc.) affect ridership.
* Investigate what affect bike supply (electric vs classic), weather, time of day, etc.  has on rider selection.
  
  
  
  
  
  
  
  